Constructive forcing

### 2023-09-15 slides-herrsching2023.tex
Beginning of notes slides

@@

{\usebackgroundtemplate{\begin{minipage}{\paperwidth}\centering\includegraphics[height=\paperheight]{multiverse-faded-169}\end{minipage}}
\begin{frame}[c]
  \centering
  \color{white}

@@

  \bigskip
  \bigskip
  \bigskip
  \bigskip
  \bigskip
  \bigskip
  \bigskip

%  \colorbox{black!50}{Autumn school on} \\
%  \colorbox{black!50}{\emph{Proof and Computation}} \\
%  \colorbox{black!50}{in Herrsching} \\
%  \ \\
%  \colorbox{black!50}{September 20th to September 16th, 2023}
%
%  \colorbox{black!50}{Ingo Blechschmidt} \\
%  \par
\end{frame}}

@@

\begin{frame}{}
  \begin{center}\includegraphics[width=0.2\textwidth]{eigenvector}\end{center}

@@

% \begin{document}

\newcommand\ytl[2]{%
  \parbox[b]{4.5em}{\hfill{#1}~$\cdots\cdots$~}%
  \makebox[0pt][c]{\color{mypurple}$\bullet$}{\color{mypurple}\vrule}\quad%
  \parbox[c]{9.3cm}{\vspace{7pt}\raggedright#2\\[7pt]}%
  \\[-2.0pt]}
\begin{frame}{A brief timeline}
  \begin{columns}[t]
    \begin{column}{0.80\textwidth}
      \ytl{1878}{Cantor advances the \hil{continuum hypothesis}, the claim
      that~$2^{\aleph_0} = \aleph_1$.}
      \pause
      \ytl{1910s}{Zermelo--Fraenkel set theory emerges.}
      \pause
      \ytl{1920s}{Set theorists pursue additional axioms to
      settle~\textsc{ch} \\ (one way or another).}
      \pause
      \ytl{1938}{Gödel proves: If~\textsc{zfc} is consistent, so
      is~\textsc{zfc}+\textsc{ch}.}
      \pause
      \ytl{1963}{Cohen proves: If~\textsc{zfc} is consistent, so
      is~\textsc{zfc}+$\neg$\textsc{ch}.}
      \pause
      \ytl{2011}{Hamkins offers his paper on the \hil{multiverse position} in
      the philosophy of set theory.}
      \pause
      %arguing that the program of pursuing
      %additional axioms (while successful in many ways) is doomed to fail
      %to settle~\textsc{ch}.}
      \ytl{2016}{Oldenziel proposes to study the modal multiverse of parametrized
      mathematics.}
      \parbox[b]{4.5em}{\hfill\phantom{x}}\makebox[0pt][c]{\phantom{b}}{\color{mypurple}\vrule}\\[-6.5pt]
      \parbox[b]{4.5em}{\hfill\phantom{x}}\makebox[0pt][c]{\phantom{b}}{\color{mypurple}\vrule}\\[-6.5pt]
      \parbox[b]{4.5em}{\hfill\phantom{x}}\makebox[0pt][c]{\phantom{b}}{\color{mypurple}\vrule}\\[-6.5pt]
    \end{column}

>>>>

\setbeameroption{show notes on second screen=bottom}
\newcommand{\jnote}[2]{\only<#1>{\note{\setlength\parskip{\medskipamount}\footnotesize\justifying#2\par}}}

@@

{\usebackgroundtemplate{\begin{minipage}{\paperwidth}\centering\includegraphics[height=\paperheight]{multiverse-faded-169}\end{minipage}}
\begin{frame}[c]
  \jnote{1}{These are the slides for a three-part lecture series for the autumn school
  \fixedhref{https://www.mathematik.uni-muenchen.de/~schwicht/pc23.php}{\emph{Proof and Computation}}
  organized by Klaus Mainzer, Peter Schuster and Helmut Schwichtenberg, held in
  Herrsching from September 20th to September 16th, 2023.}

  \centering
  \color{white}

@@

  \bigskip
  \bigskip
  \bigskip
  \bigskip
  \bigskip
  \bigskip
  \bigskip
\end{frame}}

@@

\begin{frame}{}
  \jnote{3}{
    The slide presents a question on calculus. Constructive forcing
    establishes a connection with constructive linear algebra; we will understand
    that:
    \begin{enumerate}\justifying
      \item The eigenvalues (defined as the zeros of the characteristic
      polynomial) locally depend continuously on the parameter \hil{because}
      it is a theorem of constructive linear algebra that symmetric matrices have
      a full list of eigenvalues.

      \item The eigenvectors can \hil{not} be expected to locally depend
      continuously \hil{because} it is \hil{not} a theorem of constructive
      linear algebra that symmetric matrices have an eigenvector basis.

      \item There is a dense open subset of the parameter space which restores
      continuous dependence \hil{because} it \hil{is} a theorem of constructive
      linear algebra that every symmetric matrix does \emph{not~not} have an
      eigenvector basis.
    \end{enumerate}

    A simple example where the eigenvectors cannot be chosen to locally depend
    continuously on the parameter is
    \fixedhref{https://mathoverflow.net/a/60563/31233}{recorded here}. Removing
    the origin there yields a suitable dense open subset.
  }

  \begin{center}\includegraphics[width=0.2\textwidth]{eigenvector}\end{center}

@@

\newcommand\ytl[2]{%
  \parbox[b]{4.5em}{\hfill{#1}~$\cdots\cdots$~}%
  \makebox[0pt][c]{\color{mypurple}$\bullet$}{\color{mypurple}\vrule}\quad%
  \parbox[c]{9.3cm}{\vspace{7pt}\raggedright#2\\[7pt]}%
  \\[-2.0pt]}
\begin{frame}{A brief timeline}
  \jnote{4}{
    Gödel's proof is by the \hil{$L$-translation}, where $L$ is the
    ``constructible universe''. This translation ``relativizes quantification
    to~$L$'', for instance the~$L$-translation of
    \[ \varphi \defeqv \bigl(\forall x\_ \exists y\_ \ldots\bigr) \qquad\text{is}\qquad
    \varphi^L \equiv \bigl(\forall(x \in L)\_ \exists(y \in L)\_ (\ldots)^L\bigr). \]
    We can then verify, in a weak metatheory such as~\textsc{pra}, that for
    every formula~$\varphi$ in the language of set theory,
    \[ \text{if $\textsc{zfc} \vdash \varphi$, then $\textsc{zf} \vdash
    \varphi^L$}. \]
    Specializing to~$\varphi \defeqv \bot$ we obtain: If~\textsc{zfc} is
    inconsistent, then so is~\textsc{zf}. The axiom of choice does not
    introduce new inconsistencies.

    In modern semantic language: While the axiom of choice might fail in the
    base universe~$V$ (= the class of all sets), it always holds in~$L$.
    Gödel's~$L$ was the first \emph{inner model} (= class-sized model of set
    theory) explicitly studied, nowadays we know many.
  }

  \jnote{5}{
    For his proof, Cohen invented the technique of \emph{forcing}, situated in
    classical mathematics where the base universe~$V$ is assumed to validate
    the axioms of~\textsc{zfc}.

    Just as a given ring~$R$ or group can be extended in various ways, to include
    ``generic elements'' as in~$R[X]$ or elements with prescribed relations as
    in~$R[X]/(X^2+1) =\vcentcolon R[i]$. The idea of forcing is to construct
    similar such extensions, but of universes (traditionally set-sized models
    of~\textsc{zfc}, but also class-sized models of~\textsc{zfc}, or models
    of intuitionistic set theories, or models of type theories, or even models
    of arithmetic) instead of rings.

    In semantic language, from a high level the idea is the following: Whether
    the base universe~$V$ contains a cardinal number intermediate
    between~$\aleph_0$ and~$2^{\aleph_0}$ is uncertain. But there is a certain
    extension of the base universe---constructed by forcing---which does
    contain such a number. Like the base~$V$, this forcing extension still
    validates the axioms of~\textsc{zfc}. Hence there cannot be a
    \textsc{zfc}-proof of~\textsc{ch}, as in Cohen's extension~$\neg\textsc{ch}$
    holds.

    Syntactically, Cohen's forcing provides us with an explicit formula
    translation~$\varphi \mapsto \varphi^C$ such that~\textsc{pra} proves:
    For every formula~$\varphi$, if~$\textsc{zfc}{\text{+}}\neg\textsc{ch} \vdash \varphi$
    then~$\textsc{zfc} \vdash \varphi^C$.
    \bigskip
  }

  \begin{columns}[t]
    \begin{column}{0.80\textwidth}
      \ytl{1878}{Cantor advances the \hil{continuum hypothesis}, the claim
      that~$2^{\aleph_0} = \aleph_1$.}
      \pause
      \ytl{1910s}{Zermelo--Fraenkel set theory emerges.}
      \pause
      \ytl{1920s}{Set theorists pursue additional axioms to
      settle~\textsc{ch} \\ (one way or another).}
      \pause
      \ytl{1938}{Gödel proves: If~\textsc{zfc} is consistent, so
      is~\textsc{zfc}+\textsc{ch}.}
      \pause
      \ytl{1963}{Cohen proves: If~\textsc{zfc} is consistent, so
      is~\textsc{zfc}+$\neg$\textsc{ch}.}
      \pause
      \ytl{2011}{Hamkins offers his paper on the \hil{multiverse position} in
      the philosophy of set theory.}
      \pause
      %arguing that the program of pursuing
      %additional axioms (while successful in many ways) is doomed to fail
      %to settle~\textsc{ch}.}
      \ytl{2016}{Oldenziel proposes to study the modal multiverse of parametrized
      mathematics.}
      \parbox[b]{4.5em}{\hfill\phantom{x}}\makebox[0pt][c]{\phantom{b}}{\color{mypurple}\vrule}\\[-6.5pt]
      \parbox[b]{4.5em}{\hfill\phantom{x}}\makebox[0pt][c]{\phantom{b}}{\color{mypurple}\vrule}\\[-6.5pt]
      \parbox[b]{4.5em}{\hfill\phantom{x}}\makebox[0pt][c]{\phantom{b}}{\color{mypurple}\vrule}\\[-6.5pt]
    \end{column}
